---
title: "Simplifying Genomic Annotations in R"
author: "Mohammed Omar Elsiddieg Abdallah"
date: "`r doc_date()`"
package: "`r pkg_ver('BiocStyle')`"
#  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{"Simplifying Genomic Annotations in R"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
output: 
   BiocStyle::html_document
---
# Introduction to CellbaseR
This R package makes use of the exhaustive RESTful Web service API that has been
implemented for the Cellabase database. It enable researchers to query and 
obtain a wealth of biological information from a single database saving a lot 
of time. Another benefit is that researchers can easily make  queries about 
different biological topics and link all this information together as all 
information is integrated.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# CellbaseR Classes and Methods
## The CellbaseR class  
This is an S4 class that holds the basic configuration for connecting to the 
Cellbase web services. Let us start by loading the library. 

```{r , eval=FALSE, message=FALSE}
# to get the default CellbaseR object (human data, from genome GRCh37)
library(cellbaseR)
# A default cellbaseR object is created as follows
cb <- CellBaseR()
# to change the default species from human to mouse for example
mm <-CellBaseR(species = "mmsculus") 

```

You can see the avaiable species and their corresonding data by
running the cbSpecies method passing  the CellbBseR object you just created 

```{r, message=FALSE, warning=FALSE}
library(cellbaseR)
cb <- CellBaseR()
sp <- cbSpeciesClient(object = cb)
# This will give you a CellbaseResult object
# to get the dataframe of all available species
sp <- cbData(sp)
str(sp,1)

```

## The CellbaseR Methods
In addition to the cbSpecies you have seen already, The cellbaseR package
provide many methods to query the cellbase webservices, they include:  
 
-  cbGeneClient  
-  cbSnpClient  
-  cbProteinClient   
-  cbTranscriptClient 
-  cbRegionClient 
-  cbVariantClient   
-  cbXrefClient   
-  cbClinicalClient 

In all cases the user is expected to provide the ids for the query and the 
resource to be queried, in addition to the CellbaseQuery object they created.
For example, a query through the cbGene will look something like this

### cbGeneClient
```{r, message=FALSE, warning=FALSE}
library(cellbaseR)
cb <- CellBaseR()
genes <- c("TP73","TET1")

res <- cbGeneClient(object = cb, ids = genes, resource = "info")
# to get the resulting data.frame run cellbaseData()
res <- cbData(object = res)
str(res,2)
# as you can see the res dataframe also contains a transcriots column 
# which is in fact a list column of nested dataframes, to get the
# trasnscripts data.frame for first gene
TET1_transcripts <- res$transcripts[[1]]
str(TET1_transcripts,1)
```


### cbRegionClient  
```{r, message=FALSE, warning=FALSE}
# making a query through cbRegion to get all the clinically relevant variants 
# in a specific region
library(cellbaseR)
cb <- CellBaseR()
res <- cbRegionClient(object=cb,ids="17:1000000-1100000",
resource="clinical")
# to get all conservation data in this region
res <- cbRegionClient(object=cb,ids="17:1000000-1100000",
resource="conservation")
#likewise to get all the regulatory data for the same region
res <- cbRegionClient(object=cb,ids="17:1000000-1100000", resource="regulatory")
res <- cbData(res)
str(res,1)
```
### cbVraiantClient 
Getting annotation for a specific variant
```{r, eval=FALSE,message=FALSE, warning=FALSE}
library(cellbaseR)
cb <- CellBaseR()
res2 <- cbVariantClient(object=cb, ids="1:169549811:A:G", resource="annotation")
# to get the data 
res2 <- cbData(res2)
str(res2, 1)
```

A very powerfull feature of cellbaseR is ability to fetch a wealth of clinical 
data, as well as abiliy to fiter these data by gene, phenotype, rs, etc…

### cbClinical

```{r, eval=TRUE, message=FALSE, warning=FALSE}
library(cellbaseR)
cb <- CellBaseR()
# First we have to specify aour filters, we do that by creating an object of
# class CellbaseParam
cbparam <- CellBaseParam(gene="TET1", genome="GRCh38")
cbparam
# Note that cbClinical does NOT require any Ids to be passed, only the filters
# and of course the CellbaseQuery object
res <- cbClinicalClient(object=cb,filters=cbparam)
res
res <- cbData(res)
str(res,1)

```

# CellbaseR utilities
## CreateGeneModel 
A usefull utility for fast building of gene models, compatible with Gviz 
package for genomic visualization

```{r,message=FALSE, warning=FALSE, eval=FALSE}
library(cellbaseR)
cb <- CellBaseR()
test <- createGeneModel(object = cb, region = "17:1500000-1550000")
library(Gviz)
testTrack <- GeneRegionTrack(test)
plotTracks(testTrack, transcriptAnnotation='symbol')
# then you can use this track in conjunction with any other tracks
ideoTrack <- IdeogramTrack(genome = "hg19", chromosome = "chr17")
axisTrack <- GenomeAxisTrack()
from <- min(test$start)-5000
to <- max(test$end)+5000
plotTracks(list(ideoTrack,axisTrack,testTrack),from = from, to = to, transcriptAnnotation='symbol')


```

## cbAnnotateVcf  
This utility allows users to annoate genomic variants from small to medium-sized vcf files directly from the cellbase server, with a wealth of genomic annotations including riche clinical annotations like clinvar, gwas, and cosmic data, as well as conservation and functional scores like phast and cadd  scores, without the need to download any files.  
```{r, message=FALSE, warning=FALSE, eval=FALSE}
library(cellbaseR)
cb <- CellBaseR()
fl <- system.file("extdata", "chr22.vcf.gz", package="VariantAnnotation")
res <- cbAnnotateVcf(object=cb, file=fl)
res <- cbData(res)
res

```





